Understanding meta-analysis through data simulation with applications to power analysis

### 2023-07-24 documents/paper/paper.tex
update

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{meta-analysis, monte carlo simulations, power analysis\newline\indent Word count: 7877}
\usepackage{csquotes}
\usepackage{setspace}
\captionsetup[figure]{font={stretch=1,footnotesize}}
\usepackage{float}
\usepackage{orcidlink}
\newcommand{\rev}[1]{\textcolor{red}{#1}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Understanding meta-analysis through data simulation with applications to power analysis},
  pdfauthor={Filippo Gambarota 1 \& Gianmarco Altoè 1},
  pdflang={en-EN},
  pdfkeywords={meta-analysis, monte carlo simulations, power analysis},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

A meta-analysis is an essential tool for combining knowledge from multiple studies quantitatively. Meta-analysis is commonly used together with a systematic review of the literature. The meta-analysis has several advantages. Firstly, it allows combining evidence from multiple studies assigning more weight to studies with lower estimation variability. \color{red} Then using meta-regression, it is possible to include variables (i.e., moderators) to explain the observed heterogeneity (Borenstein, Hedges, Higgins, \& Rothstein, 2009, pp. 187--203). More recently, location-scale models have been developed to include predictors also on the residual heterogeneity (i.e., Viechtbauer \& López-López, 2022). \color{black} Finally, considering the replication crisis, there are statistical methods to determine the presence and extent of the publication bias. Despite the advantages, meta-analysis implementation is not always straightforward, especially for complex data structures. Additionally, the fact that there are several introductory and advanced resources to understand meta-analysis (Borenstein et al., 2009; Harrer, Cuijpers, Furukawa, \& Ebert, 2021; Schmid, Stijnen, \& White, 2022), to our knowledge, there are no introductory resources about how to simulate realistic meta-analytic data. Simulating data has several advantages because it requires understanding the statistical method and the data-generation process. Furthermore, data simulation is the primary tool when it comes to evaluating a new analysis method, estimating the statistical power, or understanding the long-run behavior of our data generation process (Gelman \& Hill, 2006, pp. 155--176; Gelman, Hill, \& Vehtari, 2020, pp. 69--76; Ingalls, 2011). A recent paper by DeBruine and Barr (2021), which deeply inspired the current work, proposed a stimulating way to understand linear mixed effects via data simulation. Simulating data is also a powerful educational tool within this framework. For these reasons, this work aims to introduce the basic concepts of meta-analysis and Monte Carlo simulations for \textcolor{red}{equal} and random-effects, and meta-regression models with applications also for statistical power calculation. In the first section, we will introduce basic concepts of the meta-analysis that are useful for setting up the simulation. We will evaluate the effect size, variance calculation, and the \textcolor{red}{equal} vs.~random effects model distinction. Then we will describe how to simulate data for these models and simulate a meta-regression model with a categorical and numerical predictor. Finally, we will introduce the power analysis extending the previous examples to estimate the statistical power. We used the R statistical programming language (version 4.2.3; R Core Team, 2023). We assume the reader is familiar with the basic concepts of R, but core functions will be explained. Code and materials are available on the OSF repository (\url{https://osf.io/54djn/}). \color{red} A theoretical introduction, simulation examples for multivariate and multilevel models, and more details about the coding approach are available in the supplementary materials. \color{black}

@@

The first step of a meta-analysis is to extract information from included studies. This common measure should give an immediate idea of the direction (i.e., the treatment improves or reduces performance) and the size of the effect. Standardized effect size measures (see Lakens, 2013 for an overview), such as the Standardized Mean Difference (SMD), which could be estimated by Cohen's \(d\) (Cohen, 1988) or the Pearson correlation coefficient \(\rho\) which could be calculated using the associated sample estimator \emph{r} are commonly used to compare heterogeneous outcome variables\footnote{\color{red} When using bounded effect size indexes such as the Pearson correlation, Fisher's \emph{z} transformation is commonly used (Borenstein et al., 2009, pp. 41--43) \color{black}}. If all studies used the same raw measure, such as reaction times, it is possible to directly meta-analyze the studies without standardizing e.g.~using the unstandardized mean difference (UMD, Borenstein et al., 2009, pp. 21--24).

@@

The core of a meta-analysis is combining the results of multiple studies giving more weight to studies that provide a more precise effect estimation. We can find essentially three meta-analysis models: the equal-effects, the fixed-effects and the random-effects model (Hedges \& Vevea, 1998; Laird \& Mosteller, 1990). The \emph{equal-effects} model assume that each study included in the meta-analysis is a more or less precise estimation of the true underlying effect (\(\theta\)). In other terms, we are assuming that there is no variability (i.e., heterogeneity) among true effect sizes. On the other side, suppose some study-level characteristics (e.g., participants' age, sex, or socioeconomic status) or the experimental paradigm (e.g., type of memory task or difficulty) could impact the treatment effect. In this case we have true variability (i.e., heterogeneity) among studies. The \emph{random-effects} model assume a distribution of real effects with mean \(\mu_{\theta}\) and variance \(\tau^{2}\) thus estimating the heterogeneity. Finally, the \emph{fixed-effects} estimate the average effect of the included pool of studies ignoring the presence of heterogeneity. While the distinction between the equal and fixed-effects model is theoretical, the estimated model is the exactly the same\footnote{See \href{https://wviechtb.github.io/metafor/reference/misc-models.html}{wviechtb.github.io/metafor/reference/misc-models.html} for a detailed explanation}. The \emph{random-effects} model assume and estimate heterogeneity among effect sizes leading to different model parameters. From an inferential point of view, the \emph{random-effects} model provide unconditional inference to the population of effect sizes while the \emph{fixed-effects} model estimate the average effect of the selected studies, and not population-level parameters. The \emph{equal-effects} model assume a single true population effect to be estimated. The Figure \ref{fig:img-equal-vs-random} depicts the theoretical distinction between the equal and random-effects model. In the presence of heterogeneity, we can include moderators (e.g., the type of experimental setup) into a meta-regression model to explain the effect sizes heterogeneity. \color{red} An interesting proposal to combine evidence from these different models into a single analysis is called Bayesian model-averaged meta-analysis (Berkhout, Haaf, Gronau, Heck, \& Wagenmakers, 2023; Gronau, Heck, Berkhout, Haaf, \& Wagenmakers, 2021). \color{black}

@@

\caption{\color{red} The difference between the assumptions of the \emph{equal} and \emph{random-effects model}. Each distribution depicts the sampling distribution of \(k = 4\) hypothetical studies (\(i = 1\), \(2\), \ldots, \(4\)) with a certain observed effect size \(y_{i}\) (red squares), sampling variability \(\sigma_{\epsilon_i}^{2}\) and the 95\% confidence interval (black segment). The \emph{equal-effects} plot on the left suggests that each observed effect size has the same underlying true effect \(\theta\) (each distribution has the same mean) with a different degree of precision (e.g., Study 1 is more precise than Study 3). In practice, each study has a different observed effect size where studies with high precision (i.e., narrow sampling distributions) will be close to the real effect (\(\theta\)). The \emph{random-effects model} on the right suggests that beyond the error term (\(\epsilon_{i}\)), each real effect size is composed of a fixed part (now \(\mu_{\theta}\)) and a random part (\(\theta_{i}\)) sampled from a normal distribution with mean zero and variance \(\tau^{2}\). When \(\tau^{2}\) is zero, the random-effects reduces to a equal-effects model. \color{black}}\label{fig:img-equal-vs-random}
\end{figure}

@@

This tutorial use several R packages for the simulations, meta-analysis fitting and figures/tables. For the data manipulation we used the \emph{tidyverse} (Wickham, 2023) package. For the models fitting we used the \emph{metafor} (Viechtbauer, 2010) package. For figures and tables the \emph{ggplot2} (Wickham et al., 2023) or the \texttt{metafor::forest()} function, \emph{kableExtra} (Zhu, 2021) and \emph{papaja} (Aust \& Barth, 2022) packages. We set the \texttt{seed} for reproducibility of the simulation environment.
\color{black}

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(dplyr) }\CommentTok{\# for data manipulation (within tidyverse)}
\FunctionTok{library}\NormalTok{(tidyr) }\CommentTok{\# for data manipulation (within tidyverse)}
\FunctionTok{library}\NormalTok{(metafor) }\CommentTok{\# for meta{-}analysis models fitting}
\FunctionTok{library}\NormalTok{(purrr) }\CommentTok{\# for *apply like functions}
\NormalTok{seed }\OtherTok{\textless{}{-}} \DecValTok{2023} \CommentTok{\# general seed for all simulations}
\end{Highlighting}
\end{Shaded}

@@

Using the \texttt{sim\_study()} function multiple times (with the appropriate adjustments) we can generate a series of studies simulating a data set for a meta-analysis (using the \texttt{sim\_studies()} function). After each example, we will compute the appropriate model (e.g., \emph{equal} or \emph{random-effects}) using the \texttt{metafor} package (Viechtbauer, 2010) to check the recovery of simulated parameters\footnote{\color{red} By default \texttt{metafor} use the Wald \emph{z} test on model parameters but other inference methods are available e.g.~the Knapp and Hartung (2003) method by setting \texttt{test="knha"} within the \texttt{rma()} function. This method is useful especially when the number of studies is low (see \url{https://wviechtb.github.io/metafor/reference/misc-recs.html} for more details) \color{black}}. Table \ref{tab:notation-tab} summarizes the simulation parameters notation in equations and code.

@@

\caption{\color{red} Forest plots of two simulated \emph{equal-effects models}. On the left, the simulated model has \(nt,nc = 30\) for each included study while on the right the sample size for each study is \(nt,nc = 500\). Given that the data were generated under a \emph{equal-effects model}, when the sample size is high (on the right) each study is aligned on the real effect size because the error component (\(\epsilon_{i}\)) is almost entirely reduced. The average effect size is similar (depending on the random numbers generation) between the two scenarios while the estimation precision (the width of the black diamond) is narrower on the right. \color{black}}\label{fig:equal-effects-high-low-precision}
\end{figure}

@@

Compared to the \emph{equal-effects model} we need to generate another adjustment to the overall effect \(\mu_{\theta}\) from a normal distribution with mean 0 and variance \(\tau^{2}\). The real effect size for the study \(i\) will be \(\mu_{\theta} + \theta_{i}\) where \(\theta_{i}\) are the random-effects regulated by the \(\tau^{2}\) parameter.

@@

An important aspect of the \emph{random-effects model} is the interplay between the heterogeneity (\(\tau^{2}\)) and the sampling variability (\(\sigma_{i}^{2}\)). As the number of studies \(k\) increases, the estimation of \(\mu_{\theta}\) and \(\tau^{2}\) becomes more precise \color{red} (Blázquez-Rincón, Sánchez-Meca, Botella, \& Suero, 2023; Rubio-Aparicio et al., 2018) \color{black}. Additionally, as the sample size of each study decreases, each \(\theta_{i}\) will be estimated with higher precision but as long as \(\tau^{2} \neq 0\) there will be variability among effect sizes. In other words, increasing the sample size of each study or the number of studies will not affect the value of \(\tau^{2}\) but only the estimation precision (see Borenstein et al., 2009, Chapter 16, Figure 16.6 for a clear explanation). This can be easily demonstrated using the previous simulation, increasing each study sample size. Figure \ref{fig:random-effects-high-low-precision} depicts the same meta-analysis but with different precision in estimating the study \(y_{i}\). Compared to Figure \ref{fig:equal-effects-high-low-precision}, increasing the sample size of primary studies improves the estimation of each study without reducing the between-studies heterogeneity.

@@

\begin{align}
\begin{gathered}Random-effects model true effect
\tilde{v} = \frac{(k-1) \sum w_i}{(\sum w_i)^2 - \sum w_i^2}
\label{eq:typical-v}
\end{gathered}
\end{align}
\color{black}

From Equation \eqref{eq:isquared} and Figure \ref{fig:random-effects-high-low-precision} is clear that if each included study has a considerable sample size the total variability will be reduced, mainly driven by real heterogeneity (\(\tau^2\)). This is the crucial difference between the \textcolor{red}{equal-effects} and the random-effects model (see also Borenstein et al., 2009, pp. 117--122).

@@

Given the interpretation of \(I^{2}\) it is possible to simulate a meta-analysis fixing a certain \(I^{2}\) value. The only caveat is fixing the \(\tilde{v}\). When the sample size of each study is the same\footnote{The homogeneous sample size of primary studies assumption is commonly used to calculate the statistical power (see Borenstein et al., 2009, Chapter 29)}, \(\tilde{v} = \sigma_{\epsilon_i}^{2}\). In the other case, \(\tilde{v}\) needs to be calculated from sampling variances as reported in Higgins and Thompson (2002) and cannot be easily fixed a priori. With the assumption of homogeneous sample size across studies\footnote{In case of heterogeneous sample sizes generated from a probability distribution (e.g., Gaussian) the \emph{typical} sampling variance calculated using the expected value of the distribution can be used to somehow fix the (average) \(I^2\) (see the supplementary materials for a simulated example)}, we can solve Equation \eqref{eq:isquared} for \(\tau^{2}\) obtaining the heterogeneity value associated with a certain \(I^{2}\) as reported in Equation \eqref{eq:tau2-from-isquared}.
\color{black}

@@

A common example of meta-regression is by including a categorical predictor, including information about study-level features. In our example, a group of studies uses an online memory task while others use a standard lab-based task. The Equation \eqref{eq:random-effects-model1} can be easily extended for a meta-regression model by including a variable encoding the type of task (online vs.~lab-based) and the expected difference between the two levels of the moderator (i.e., the \emph{lab vs.~online effect}). In regression terms (see equation \eqref{eq:meta-regression}), we could use a dummy variable (\(X_{1}\)) that takes the value of 0 for the \emph{lab-based task} (\emph{L}) and a value of 1 for an \emph{online task} (\emph{O})\footnote{The \emph{dummy coding} (also known as treatment coding) is the default in R and \texttt{metafor} but other coding schemes could be used (see Schad, Vasishth, Hohenstein, \& Kliegl, 2020 for an overview of contrast coding schemes). \color{red} As in standard linear regression, the contrast coding of categorical factors or the presence/absence of the intercept influence the interpretation of estimated model parameters and omnibus meta-regression tests (see \url{https://www.metafor-project.org/doku.php/tips:models_with_or_without_intercept} for a detailed discussion)} \color{black}. Now we fix \(\beta_{1}\) to be the \emph{lab vs.~online effect} (i.e., the expected mean difference between the two groups of studies) and the product between \color{red}\(\beta_{1}X_{1_i}\) \color{black} will consider the \emph{lab vs.~online effect}. Crucially, despite \(\tau^{2}\) is still the heterogeneity between effect sizes, now we need to fix \(\tau^{2}\) considering that we included a moderator. In other terms, \color{red} \(\delta_i \sim \mathcal{N}(0, \tau_{r}^{2})\) \color{black} where \(\tau^2_r\) is the residual heterogeneity after including the moderator. We can describe our model using equation \eqref{eq:meta-regression-equations} according to the value of \(X_{1_i}\) (L = lab-based experiments, O = online experiments).

@@

Now we can fit the meta-regression model with the \texttt{rma} function as for the random-effects model with the addition of \texttt{mods\ =\ \textasciitilde{}\ exp} that indicates which variable/s to consider as moderators. The results are presented in Table \ref{tab:res-meta-reg-dummy}. Now the model will estimate an \emph{intercept} parameter (i.e., \(\beta_0\)) that is the value of \(y\) when \(X_1\) is zero (i.e., for lab-based studies) or in other terms the expected value for lab-based studies. Then the \(\beta_1\) parameter represents the estimated difference in \(y\) between the values of \(X_1\) (i.e., lab-based vs online experiments). As said before, \color{red} \(\tau^2_r\) \color{black} is now the residual heterogeneity that is interpreted as the variability between effect sizes after controlling for the moderator \(X_1\).

@@

\caption{Meta-regression results for the \emph{random-effects model} with a numerical moderator. Each effect size is represented with a black dot where the dimension represents the weight according to the inverse of the precision. The line represents the estimated meta-regression slope with the 95\% confidence interval (grey bands).}\label{fig:plot-meta-reg-numerical-r2}
\end{figure}

@@

The previous simulation examples can be easily implemented for multiple purposes. For example, we can use different effect sizes and variance estimators when using the \texttt{sim\_study()} function to check the impact on the fitted meta-analysis model. However, one of the most critical applications is estimating the power of a specific statistical model. \color{red} One of the purpose of power analysis by simulations is estimating the required number of studies to detect an hypothetical effect size when planning a meta-analysis. At the same time, the meta-analysis that we presented (\emph{two-level} equal or random-effects model) can be considered as a multi-lab (e.g., Klein et al., 2018) study where experiments are planned and not collected from the literature. In fact, we can use the same simulation approach to optimize the number of participants and studies when planning a multi-lab project. \color{black}

As explained in the introduction, there are several approaches and tools to estimate the power of \emph{equal} and \emph{random-effects models} (see Borenstein et al., 2009, Chapter 29; Harrer, Cuijpers, \& Ebert, 2019, Chapter 14). These methods are easy to implement but made strong assumptions, such as the homogeneity of sample size, and did not consider the uncertainty in estimating \(\tau^{2}\). Jackson and colleagues (2017) partially solve the issue by developing an interesting method that takes into account the uncertainty in estimating \(\tau^{2}\) without using simulations. However, only Monte Carlo simulations can \textcolor{red}{handle} complex scenarios and extra parameters.

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{meta-analysis, monte carlo simulations, power analysis\newline\indent Word count: 7782}
\usepackage{csquotes}
\usepackage{setspace}
\captionsetup[figure]{font={stretch=1,footnotesize}}
\usepackage{float}
\usepackage{orcidlink}
\newcommand{\rev}[1]{\textcolor{red}{#1}}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Understanding meta-analysis through data simulation with applications to power analysis},
  pdfauthor={Filippo Gambarota 1 \& Gianmarco Altoè 1},
  pdflang={en-EN},
  pdfkeywords={meta-analysis, monte carlo simulations, power analysis},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

A meta-analysis is an essential tool for combining knowledge from multiple studies quantitatively. Meta-analysis is commonly used together with a systematic review of the literature. The meta-analysis has several advantages. Firstly, it allows combining evidence from multiple studies assigning more weight to studies with lower estimation variability. \color{red} Then using meta-regression, it is possible to include variables (i.e., moderators) to explain the observed heterogeneity (Borenstein, Hedges, Higgins, \& Rothstein, 2009, pp. 187--203). More recently, location-scale models have been developed to include predictors also on the residual heterogeneity (i.e., Viechtbauer \& López-López, 2022). \color{black} Finally, considering the replication crisis, there are statistical methods to determine the presence and extent of the publication bias. Despite the advantages, meta-analysis implementation is not always straightforward, especially for complex data structures. Additionally, the fact that there are several introductory and advanced resources to understand meta-analysis (Borenstein et al., 2009; Harrer, Cuijpers, Furukawa, \& Ebert, 2021; Schmid, Stijnen, \& White, 2022), to our knowledge, there are no introductory resources about how to simulate realistic meta-analytic data. Simulating data has several advantages because it requires understanding the statistical method and the data-generation process. Furthermore, data simulation is the primary tool when it comes to evaluating a new analysis method, estimating the statistical power, or understanding the long-run behavior of our data generation process (Gelman \& Hill, 2006, pp. 155--176; Gelman, Hill, \& Vehtari, 2020, pp. 69--76; Ingalls, 2011). A recent paper by DeBruine and Barr (2021), which deeply inspired the current work, proposed a stimulating way to understand linear \textcolor{red}{mixed-effects models} via data simulation. Simulating data is also a powerful educational tool within this framework. For these reasons, this work aims to introduce the basic concepts of meta-analysis and Monte Carlo simulations for \textcolor{red}{equal-effects}, random-effects, and meta-regression models with applications also to statistical power calculation. In the first section, we will introduce basic concepts of the meta-analysis that are useful for setting up the simulation. We will evaluate the effect size, variance calculation, and the \textcolor{red}{equal} vs.~random effects model distinction. Then we will describe how to simulate data for these models and simulate a meta-regression with categorical and numerical predictors. Finally, we will introduce the power analysis extending the previous examples to estimate the statistical power. We used the R statistical programming language (version 4.2.3; R Core Team, 2023). We assume the reader is familiar with the basic concepts of R, but core functions will be explained. Code and materials are available on the OSF repository (\url{https://osf.io/54djn/}). \color{red} A theoretical introduction, simulation examples for multivariate and multilevel models, and more details about the coding approach are available in the supplementary materials. \color{black}

@@

The first step of a meta-analysis is to extract information from included studies. This common measure should give an immediate idea of the direction (i.e., the treatment improves or reduces performance) and the size of the effect. Standardized effect size measures (see Lakens, 2013 for an overview), such as the Standardized Mean Difference (SMD), which could be estimated by Cohen's \(d\) (Cohen, 1988) or the Pearson correlation coefficient \(\rho\) which could be calculated using the associated sample estimator \emph{r} are commonly used to compare heterogeneous outcome variables\footnote{\color{red} When using bounded effect sizes such as the Pearson correlation, Fisher Z-transformation is commonly used (Borenstein et al., 2009, pp. 41--43) \color{black}}. If all studies used the same raw measure, such as reaction times, it is possible to directly meta-analyze the studies without standardizing e.g.~using the unstandardized mean difference (UMD, Borenstein et al., 2009, pp. 21--24).

@@

The core of a meta-analysis is combining the results of multiple studies giving more weight to studies that provide a more precise effect estimation. We can find essentially three meta-analysis models: the equal-effects, the fixed-effects, and the random-effects model (Hedges \& Vevea, 1998; Laird \& Mosteller, 1990). The \emph{equal-effects} model assumes that each study included in the meta-analysis is a more or less precise estimation of the true underlying effect (\(\theta\)). In other terms, we are assuming that there is no variability (i.e., heterogeneity) among true effect sizes. On the other side, suppose some study-level characteristics (e.g., participants' age, sex, or socioeconomic status) or the experimental paradigm (e.g., type of memory task or difficulty) could impact the treatment effect. In this case, we have true variability (i.e., heterogeneity) among studies. The \emph{random-effects} model assumes a distribution of real effects with mean \(\mu_{\theta}\) and variance \(\tau^{2}\) thus estimating the heterogeneity. Finally, the \emph{fixed-effects} model estimate the average effect of the included pool of studies ignoring the presence of heterogeneity. While the distinction between the equal and fixed-effects model is theoretical, the estimated model is the exactly the same\footnote{See \href{https://wviechtb.github.io/metafor/reference/misc-models.html}{wviechtb.github.io/metafor/reference/misc-models.html} for a detailed explanation}. The \emph{random-effects} model assumes and estimates heterogeneity among effect sizes leading to different model parameters. From an inferential point of view, the \emph{random-effects} model provides unconditional inference to the population of effect sizes while the \emph{fixed-effects} model estimates the average effect of the selected studies and not population-level parameters. The \emph{equal-effects} model assumes a single true population effect to be estimated. Figure \ref{fig:img-equal-vs-random} depicts the theoretical distinction between the equal and random-effects model. In the presence of heterogeneity, we can include moderators (e.g., the type of experimental setup) into a meta-regression model to explain the effect size heterogeneity. \color{red} An interesting proposal to combine evidence from these different models into a single analysis is called Bayesian model-averaged meta-analysis (Berkhout, Haaf, Gronau, Heck, \& Wagenmakers, 2023; Gronau, Heck, Berkhout, Haaf, \& Wagenmakers, 2021). However, this model is beyond the scope of the tutorial. \color{black}

@@

\caption{\color{red} The difference between the assumptions of the \emph{equal} and \emph{random-effects model}. Each distribution depicts the sampling distribution of \(k = 4\) hypothetical studies (\(i = 1\), \(2\), \ldots, \(4\)) with a certain observed effect size \(y_{i}\) (red squares), sampling variability \(\sigma_{\epsilon_i}^{2}\) and the 95\% confidence interval (black segment). The \emph{equal-effects} plot on the left suggests that each observed effect size has the same underlying true effect \(\theta\) (each distribution has the same mean) with a different degree of precision (e.g., Study 1 is more precise than Study 3). In practice, each study has a different observed effect size where studies with high precision (i.e., narrow sampling distributions) will be close to the real effect (\(\theta\)). The \emph{random-effects model} on the right suggests that beyond the error term (\(\epsilon_{i}\)), each real effect size is composed of a fixed part (now \(\mu_{\theta}\)) and a random part (\(\delta_{i}\)) sampled from a normal distribution with mean zero and variance \(\tau^{2}\). When \(\tau^{2}\) is zero, the random-effects reduces to a equal-effects model. \color{black}}\label{fig:img-equal-vs-random}
\end{figure}

@@

This tutorial uses several R packages for the simulations, meta-analysis fitting, and figures/tables. For the data manipulation, we used the \emph{tidyverse} (Wickham, 2023) package. For the models fitting we used the \emph{metafor} (Viechtbauer, 2010) package. For figures and tables the \emph{ggplot2} (Wickham et al., 2023) or the \texttt{metafor::forest()} function, \emph{kableExtra} (Zhu, 2021) and \emph{papaja} (Aust \& Barth, 2022) packages. We set the \texttt{seed} for reproducibility of the simulation environment.
\color{black}

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(dplyr) }\CommentTok{\# for data manipulation (within tidyverse)}
\FunctionTok{library}\NormalTok{(tidyr) }\CommentTok{\# for data manipulation (within tidyverse)}
\FunctionTok{library}\NormalTok{(metafor) }\CommentTok{\# for meta{-}analysis models fitting}
\FunctionTok{library}\NormalTok{(purrr) }\CommentTok{\# for *apply like functions (within tidyverse)}
\NormalTok{seed }\OtherTok{\textless{}{-}} \DecValTok{2023} \CommentTok{\# general seed for all simulations}
\end{Highlighting}
\end{Shaded}

@@

Using the \texttt{sim\_study()} function multiple times (with the appropriate adjustments) we can generate a series of studies simulating a data set for a meta-analysis (using the \texttt{sim\_studies()} function). After each example, we will compute the appropriate model (e.g., \emph{equal} or \emph{random-effects}) using the \texttt{metafor} package (Viechtbauer, 2010) to check the recovery of simulated parameters\footnote{\color{red} By default \texttt{metafor} use the Wald \emph{z} test on model parameters but other inference methods are available e.g.~the Knapp and Hartung (2003) method by setting \texttt{test="knha"} within the \texttt{rma()} function. This method is useful especially when the number of studies is low (see \href{https://wviechtb.github.io/metafor/reference/misc-recs.html}{wviechtb.github.io/metafor/reference/misc-recs.html} for more details) \color{black}}. Table \ref{tab:notation-tab} \textcolor{red}{summarizes the notation used in simulations and code in equations and code.}

@@

\caption{\color{red} Forest plots of two simulated \emph{equal-effects models}. On the left, the simulated model has \(nt,nc = 30\) for each included study while on the right the sample size for each study is \(nt,nc = 500\). Given that the data were generated under a \emph{equal-effects model}, when the sample size is high (on the right) each study is aligned on the real effect size because the error component (\(\epsilon_{i}\)) \textcolor{red}{is close to zero}. The average effect size is similar (depending on the random numbers generation) between the two scenarios while the estimation precision (the width of the black diamond) is narrower on the right. \color{black}}\label{fig:equal-effects-high-low-precision}
\end{figure}

@@

Compared to the \emph{equal-effects model} we need to generate another adjustment to the overall effect \(\mu_{\theta}\) from a normal distribution with mean 0 and variance \(\tau^{2}\). The real effect size for the study \(i\) will be \(\mu_{\theta} + \delta_{i}\) where \(\delta_{i}\) are the random-effects regulated by the \(\tau^{2}\) parameter.

@@

An important aspect of the \emph{random-effects model} is the interplay between the heterogeneity (\(\tau^{2}\)) and the sampling variability (\(\sigma_{i}^{2}\)). As the number of studies \(k\) increases, the estimation of \(\mu_{\theta}\) and \(\tau^{2}\) becomes more precise \color{red} (Blázquez-Rincón, Sánchez-Meca, Botella, \& Suero, 2023; Rubio-Aparicio et al., 2018) \color{black}. Additionally, as the sample size of each study decreases, each \(\delta_{i}\) will be estimated with higher precision but as long as \(\tau^{2} \neq 0\) there will be variability among effect sizes. In other words, increasing the sample size of each study or the number of studies will not affect the value of \(\tau^{2}\) but only the estimation precision (see Borenstein et al., 2009, Chapter 16, Figure 16.6 for a clear explanation). This can be easily demonstrated using the previous simulation, increasing each study sample size. Figure \ref{fig:random-effects-high-low-precision} depicts the same meta-analysis but with different precision in estimating the study \(y_{i}\). Compared to Figure \ref{fig:equal-effects-high-low-precision}, increasing the sample size of primary studies improves the estimation of each study without reducing the between-studies heterogeneity.

@@

\begin{align}
\begin{gathered}
\tilde{v} = \frac{(k-1) \sum w_i}{(\sum w_i)^2 - \sum w_i^2}
\label{eq:typical-v}
\end{gathered}
\end{align}
\color{black}

From Equation \eqref{eq:isquared} and Figure \ref{fig:random-effects-high-low-precision} is clear that if each included study has a considerable sample size \color{red} the sampling variability (\(\tilde{v}\)) will be reduced, and the total variability will be mainly driven by real heterogeneity (\(\tau^2\)) \color{black}. This is the crucial difference between the \textcolor{red}{equal-effects} and the random-effects model (see also Borenstein et al., 2009, pp. 117--122).

@@

Given the interpretation of \(I^{2}\) it is possible to simulate a meta-analysis fixing a certain \(I^{2}\) value. The only caveat is fixing the \(\tilde{v}\). When the sample size of each study is the same\footnote{\color{red} The assumption of studies with homogeneous sample sizes is commonly used to calculate the statistical power \color{black}(see Borenstein et al., 2009, Chapter 29)}, \(\tilde{v} = \sigma_{\epsilon_i}^{2}\). In the other case, \(\tilde{v}\) needs to be calculated from sampling variances as reported in Higgins and Thompson (2002) and cannot be easily fixed a priori. With the assumption of homogeneous sample size across studies\footnote{In case of heterogeneous sample sizes generated from a probability distribution (e.g., Gaussian) the \emph{typical} sampling variance calculated using the expected value of the distribution can be used to somehow fix the (average) \(I^2\) (see the supplementary materials for a simulated example)}, we can solve Equation \eqref{eq:isquared} for \(\tau^{2}\) obtaining the heterogeneity value associated with a certain \(I^{2}\) as reported in Equation \eqref{eq:tau2-from-isquared}.
\color{black}

@@

A common example of meta-regression is by including a categorical predictor, including information about study-level features. In our example, a group of studies uses an online memory task while others use a standard lab-based task. The Equation \eqref{eq:random-effects-model1} can be easily extended for a meta-regression model by including a variable encoding the type of task (online vs.~lab-based) and the expected difference between the two levels of the moderator (i.e., the \emph{lab vs.~online effect}). In regression terms (see equation \eqref{eq:meta-regression}), we could use a dummy variable (\(X_{1}\)) that takes the value of 0 for the \emph{lab-based task} (\emph{L}) and a value of 1 for an \emph{online task} (\emph{O})\footnote{The \emph{dummy coding} (also known as treatment coding) is the default in R and \texttt{metafor} but other coding schemes could be used (see Schad, Vasishth, Hohenstein, \& Kliegl, 2020 for an overview of contrast coding schemes). \color{red} As in standard linear regression, the contrast coding of categorical factors or the presence/absence of the intercept influences the interpretation of estimated model parameters and omnibus meta-regression tests (see \url{https://www.metafor-project.org/doku.php/tips:models_with_or_without_intercept} for a detailed discussion)} \color{black}. Now we fix \(\beta_{1}\) to be the \emph{lab vs.~online effect} (i.e., the expected mean difference between the two groups of studies) and the product between \color{red}\(\beta_{1}X_{1_i}\) \color{black} will consider the \emph{lab vs.~online effect}. Crucially, despite \(\tau^{2}\) is still the heterogeneity between effect sizes, now we need to fix \(\tau^{2}\) considering that we included a moderator. In other terms, \color{red} \(\delta_i \sim \mathcal{N}(0, \tau_{r}^{2})\) \color{black} where \(\tau^2_r\) is the residual heterogeneity after including the moderator. We can describe our model using equation \eqref{eq:meta-regression-equations} according to the value of \(X_{1_i}\) (L = lab-based experiments, O = online experiments).

@@

Now we can fit the meta-regression model with the \texttt{rma} function as for the random-effects model with the addition of \texttt{mods\ =\ \textasciitilde{}\ exp} that indicates which variables to consider as moderators. The results are presented in Table \ref{tab:res-meta-reg-dummy}. Now the model will estimate an \emph{intercept} parameter (i.e., \(\beta_0\)) that is the value of \(y\) when \(X_1\) is zero (i.e., for lab-based studies) or in other terms the expected value for lab-based studies. Then the \(\beta_1\) parameter represents the estimated difference in \(y\) between the values of \(X_1\) (i.e., lab-based vs online experiments). As said before, \color{red} \(\tau^2_r\) \color{black} is now the residual heterogeneity that is interpreted as the variability between effect sizes after controlling for the moderator \(X_1\).

@@

\caption{Meta-regression results for the \emph{random-effects model} with a numerical moderator. Each effect size is represented with a black dot where the dimension represents the weight according to the inverse of the variance. The line represents the estimated meta-regression slope with the 95\% confidence interval (grey bands).}\label{fig:plot-meta-reg-numerical-r2}
\end{figure}

@@

The previous simulation examples can be easily implemented for multiple purposes. For example, we can use different effect sizes and variance estimators when using the \texttt{sim\_study()} function to check the impact on the fitted meta-analysis model. However, one of the most critical applications is estimating the power of a specific statistical model. \color{red} One of the purposes of power analysis by simulations is to estimate the required number of studies to detect an hypothetical effect size when planning a meta-analysis. At the same time, the meta-analysis that we presented (\emph{two-level} equal or random-effects model) can be considered as a multi-lab (e.g., Klein et al., 2018) study where experiments are planned and not collected from the literature. We can use the same simulation approach to optimize the number of participants and studies when planning a multi-lab project. \color{black}

As explained in the introduction, there are several approaches and tools to estimate the power of \emph{equal} and \emph{random-effects models} (see Borenstein et al., 2009, Chapter 29; Harrer, Cuijpers, \& Ebert, 2019, Chapter 14). These methods are easy to implement but made strong assumptions, such as the homogeneity of sample size, and did not consider the uncertainty in estimating \(\tau^{2}\). Jackson and colleagues (2017) partially solve the issue by developing an interesting method that takes into account the uncertainty in estimating \(\tau^{2}\) without using simulations. \textcolor{red}{However, complex simulation scenarios can be handled only using Monte Carlo methods.}
